Predicting Sequence Variability from Voronoi Tessellation of Proteins

### 2015-02-27 manuscript/voronoi/manuscript_voronoi.tex
MN update

    In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra. Figure \ref{fig:best_voronoi} (left plot) depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties. In general, we observe that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy. The cell properties are also highly associated with each other. Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in the left plot of Figure \ref{fig:best_voronoi}. On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. These cell characteristics are however, discrete (integer) quantities and in general have a limited range.

@@

		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced, using Voronoi tessellation from the field of computational geometry, that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al. (2014xx), site-specific measures of flexibility, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates. It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.

@@

    \begin{figure*}[tbh]
        \begin{center}
        \begin{tabular}{cc}  
        \includegraphics[width=3.35in]{best_voronoi_predictors_of_ER.pdf} & \includegraphics[width=3.35in]{best_voronoi_predictors_of_ER_given_area.pdf}
        \end{tabular}
        \end{center}
        \caption{{\bf Right:} A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER). Note that all cell characteristic correlate positively with ER, except sphericity which strongly negatively correlates with ER. {\bf Left:} The partial correlation of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
        \label{fig:voronoi_ER}
    \end{figure*}

>>>>

    In order to assess the prediction power site-specific variables derived from Voronoi tessellation, we first calculate the geometric centers of all side-chains for each of the proteins in the dataset and use them as the seeds of Voronoi polyhedra. Plot {\it A} of Figure \ref{fig:best_voronoi} depicts the distributions of the Spearman's correlation coefficients of five most important Voronoi cell characteristics with site-specific evolutionary rates (ER). It should be noted that all cell characteristics in the plot correlate positively with ER, except the cell sphericity which is always negatively correlated with ER and other Voronoi cell properties. In general, we observe that the cell surface area has the best prediction power compared to other cell characteristics, followed by the cell volume, cell eccentricity as defined in previous section, cell's total edge length, and the cell sphericitiy. The cell properties are also highly associated with each other. Although the Voronoi cell volume is the second best correlating variable with ER, it exhibits no significant independent correlation with ER once we control for the cell area with the median of the distribution centered at $\sim0.0$, as illustrated in plot {\it B} of Figure \ref{fig:best_voronoi}. On the contrary, the cell sphericity and eccentricity both exhibit median partial correlations of $\sim-0.1$ \& $\sim0.07$ with ER respectively, when the contribution from the Voronoi cell area is controlled. In conclusion, the cell area, volume, and edge length appear to almost represent the same property of the Voronoi cell. Other Voronoi cell characteristics, such as the number of vertices, faces and edges of the cell also tend to correlate weakly with sequence evolutionary rates. These cell characteristics are however, discrete (integer) quantities and in general have a limited range.

@@

		Throughout the previous sections, a new set of parameter-free site-specific structural quantities were introduced using Voronoi tessellation that are capable of explaining sequence evolutionary rates equally or better than many of the previously considered site-specific structural characteristics, including RSA, ddG rate as defined by Echave et al. (2014xx), site-specific measures of flexibility such as B factor, and the traditional definition of Contact Number and the Weighted Contact Number using $C_\alpha$ atomic coordinates. It is however notable that, once we recalculate WCN using the geometric center of the side chains as the representative coordinates of individual sites, the quantity WCN still outperforms all other structural quantities, including those derived from Voronoi tessellation, in explaining site-specific evolutionary rates.

@@

    \begin{figure}[tbh]
        \begin{center}
        \includegraphics[width=7in]{best_voronoi_predictors_of_ER_screen.pdf}
        \end{center}
        \caption{{\bf A:} A comparison of the prediction power of different Voronoi cell characteristics about site-specific evolutionary rates (ER). Note that all cell characteristic correlate positively with ER, except sphericity which strongly negatively correlates with ER. {\bf B:} The partial correlation of the same Voronoi cell characteristics with sequence evolutionary rates while controlling for the cell area.}
        \label{fig:voronoi_ER}
    \end{figure}
